---
title: "Practical Machine Learning Project"
author: "Frank Feng"
date: "Tuesday, June 09, 2015"
output: html_document
---
### Purpose:

The goal of this project is to build a statistic model to predict the manner ("classe" variable in the dataset) of exercise performed by a group using the measurements data collected from sensors attached to the individuals in the group.   

### Data description

The data used in this project comes from  <http://groupware.les.inf.puc-rio.br/har>. 
It recorded the measurements of Six young health participants performing one set of 10 repetitions of the Unilateral Dumbbell Biceps Curl in five different fashions: exactly according to the specification (Class A), throwing the elbows to the front (Class B), lifting the dumbbell only halfway (Class C), lowering the dumbbell only halfway (Class D) and throwing the hips to the front (Class E). The details description of data can be found in the referenced article.

### Reading the data:
After downloading the data from source link into R studio work directory, read the data file into data frames;
```{r}
pml_training<-read.csv("pml-training.csv",na.strings=c("NA",""))
pml_testing<-read.csv("pml-testing.csv",na.strings=c("NA",""))
```

### Exploring and cleaning data:
Explore the data structure,distribution, and variance: 
```{r,results='hide',warning=FALSE,message=FALSE}
summary(pml_training)
#19622 obs. 160 variable
#remove columns with NAs=19216
pml_training <- pml_training[,colSums(is.na(pml_training))<19216]
summary(pml_training)
#19622 obs. 60 variables
#check near zero variables
library(caret)
nearZeroVar(pml_training,saveMetrics=TRUE)
#check predictor types and values
str(pml_training)
```

We can see column X is a sequence number, new_window is not changing. While we are only interested to classify the movement category, the individual identify and time information should not play a role in out prediction, so we can remove column 1-6 from our training set.
```{r}
pml_training<- pml_training[,-(1:6)]
dim(pml_training)
```

### Building Model 
We can see the training data set is rather large, we will split the data set into two halves, each contains 50% of data set, one will be used for train the model, and the other will be used for cross validation and calculation of out-of-sample error. As indicated in the referenced article, we will use random forest method to classify the different category of movement (classe), and we will use k-fold cross-validation (cv) as train control options, and in this practice k=5 to save some computer time to train the model.
```{r,warning=FALSE,message=FALSE}
set.seed(201506)
inTrain <-createDataPartition(y=pml_training$classe,p=0.5,list=FALSE)
trainData <-pml_training[inTrain,]
validData <-pml_training[-inTrain,]
# use k-fold cross-validation method in train control option
trControl <- trainControl(method = "cv", number = 5,allowParallel=TRUE)
# perform training using random forest method
rfFit <- train(trainData$classe ~ ., method = "rf", trControl = trControl, data =trainData)
#final model
rfFit$finalModel
# model fit accuracy
acy<-rfFit$results$Accuracy[2]
```
From our finalModel, we obtained estimated OOB ,or out of bag (sample) error is 0.3%
and predication accuracy for the training set is `r acy`.

### Cross-Validation and out-of-sample error
Apply the model to the other half of the training set we hold out to cross validate the model and calculate the prediction accuracy and the out-of-sample error.
```{r}
# predict with the other half of the training data .
prediction_valid<-predict(rfFit,newdata=validData)
# over all accuracy
oacy<-confusionMatrix(prediction_valid,validData$classe)$overall[1]
# the out of sample error
osmpl_err<-1-oacy
```
We found overall predication accuracy for the validation set is `r oacy`  which means we did not over fit or under fit the model, or the split-ed dataset are very similar.  The out-of-sample error is `r osmpl_err` which is very close to the model estimated OOB error 0.3% from the k-fold cross-validation training control, which means the OOB error is as accurate as using a validation set of the same size as the training set.

### Test data predication
Now we can apply our model to the test set and write out answers into individual files.
```{r}
# remove NA columns
pml_testing<-pml_testing[,colSums(is.na(pml_testing))<20]
# remove first six columns
pml_testing <-pml_testing[,-(1:6)]
# predications for test data
answers <- predict(rfFit,newdata=pml_testing)
answers
#write answer to files use a function
pml_write_files = function(x){
     n = length(x)
     for(i in 1:n){
         filename = paste0("problem_id_",i,".txt")
         write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.nam=FALSE)
     }
 }
pml_write_files(answers)
```

### Reference
Velloso, E.; Bulling, A.; Gellersen, H.; Ugulino, W.; Fuks, H. Qualitative Activity Recognition of Weight Lifting Exercises. Proceedings of 4th International Conference in Cooperation with SIGCHI (Augmented Human '13) . Stuttgart, Germany: ACM SIGCHI, 2013.